---
title: "SAEM-MCMC estimation for the LFMC data"
author: ""
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc_float: true
    code-fold: true
    df-print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  cache   = TRUE
)
```

## Introduction

Leaf Fuel Moisture Content (LFMC) describes the amount of water in plants as a
percentage of dry weight. This chapter illustrates the **SAEM-MCMC** algorithm
(Stochastic Approximation EM with a Metropolis-within-Gibbs MCMC step) on
LFMC temporal dynamics for perennial grasses measured on the Eastern site of
the dataset (`leaf.type == "Grass E"`).

For each plot $i$, the LFMC is modelled with a declining four-parameter
logistic curve:

$$
\text{LFMC}_{ij} = \frac{A_i - w_i}{1 + \exp\left(\frac{m_i - t_{ij}}{s_i}\right)} + w_i + \varepsilon_{ij},
\qquad \varepsilon_{ij} \sim \mathcal N(0,\sigma^2).
$$

The individual parameters are log-transformed and gathered in
$\boldsymbol\varphi_i = (\log A_i, \log w_i, \log m_i, \log s_i)$, which are
assumed to follow a Gaussian distribution with mean $\boldsymbol\mu$ and
covariance matrix $\boldsymbol\Omega$. SAEM-MCMC is used to estimate
$\theta = (\boldsymbol\mu, \boldsymbol\Omega, \sigma^2)$.

This document relies on the helper functions in `SAEM-MCMC.R` (the
four-parameter logistic, the initialisation, the MCMC sampler, and the
SAEM-MCMC algorithm). Estimation results are cached: if the saved files
`res_SAEM_LFMC.Rdata` and `res_postSample_LFMC.Rdata` are present they are
loaded; otherwise the algorithms are run live (which takes longer).

## Packages and functions

```{r packages}
#| code-fold-summary: "Required packages"
library(ggplot2)
library(dplyr)
library(tidyr)
library(mvtnorm)
library(kableExtra)

# Figure theme
theme_set(theme_bw())

# Helper functions (four-parameter logistic, init, MCMC, SAEM-MCMC)
source("SAEM-MCMC.R")
```

## Data

```{r data}
#| code-fold-summary: "Data import and formatting"
lfmc.data <- read.table("data_lfmc.txt", header = TRUE)
lfmc.data$plot      <- as.factor(lfmc.data$plot)
lfmc.data$leaf.type <- with(lfmc.data,
  factor(leaf.type,
         levels = c("Grass W", "Grass E", "M. spinosum", "S. bracteolactus")))
levels(lfmc.data$leaf.type)[levels(lfmc.data$leaf.type) == "S. bracteolactus"] <-
  "S.  filaginoides"

# Reorder plots for consistent plotting
ord <- c("4", "5", "6", "1", "2", "3")
lfmc.data$plot_fact <- with(lfmc.data, factor(plot, levels = ord))
lfmc.data$plot      <- as.numeric(as.character(lfmc.data$plot_fact))
nPlot  <- 3
str(lfmc.data)
```

We focus on the perennial grasses of the Eastern site (`Grass E`),
which involves `r nPlot` plots.

```{r subset  }
#| code-fold: true
#| code-fold-summary: "Extract the Grass E data"
 
myData <- lfmc.data[lfmc.data$leaf.type == "Grass E", ]
```

The observed LFMC dynamics for the three Eastern grasses plots:

```{r plot-data}
#| code-fold: true
#| code-fold-summary: "Plot of the raw data"
cols <- setNames(scales::hue_pal()(6)[1:6], levels(myData$plot_fact))

ggplot(myData, aes(x = time, y = lfmc)) +
  geom_point(aes(colour = plot_fact)) +
  geom_smooth(colour = "gray", linewidth = 0.7, se = FALSE) +
  scale_colour_manual(values = cols) +
  labs(x = "Time (days)", y = "LFMC (%)", colour = "Plot")
```

## Estimation by SAEM-MCMC

### Initialisation of the algorithm 

We first build a rough guess $\boldsymbol\varphi_i^{(0)}$ of the individual
log-parameters for each plot, then derive the population parameters from
these values.

For plot $i$, `initZ` proceeds by **optimisation on the "individual curve"**:

1. $A_i$ and $w_i$ are set to the mean observed LFMC at the last and first
   observation times,
2. rearranging the logistic gives a linear model in time,
   $\log\!\big(\frac{A_i - w_i}{\text{LFMC}_{ij} - w_i} - 1\big) = \frac{m_i}{s_i} - \frac{t_{ij}}{s_i}$,
   whose fit (over all observation times of the plot) yields $m_i$ and
   $s_i$.

```{r initZ-fun}
# Optimisation-based per-individual initialisation
initZ <- function(myData) {
  nPlot <- max(myData$plot)
  Z.init <- matrix(0, nPlot, 4)
  for (i in 1:nPlot) {
    Data.i <- myData |> filter(plot == i)
    tmin.i <- min(Data.i$time)
    tmax.i <- max(Data.i$time)

    A.i <- mean(Data.i$lfmc[Data.i$time == tmax.i]) # LFMC at last time
    w.i <- mean(Data.i$lfmc[Data.i$time == tmin.i]) # LFMC at first time

    # log((A-w)/(lfmc-w)-1) ~ m/s - time/s
    U.i  <- suppressWarnings(log((A.i - w.i)/(Data.i$lfmc - w.i) - 1))
    r.i  <- which(!is.finite(U.i) | is.na(U.i))
    res_lm.i <- lm(U.i[-r.i] ~ Data.i$time[-r.i])
    s.i <- -1/res_lm.i$coefficients[2]
    m.i <- s.i * res_lm.i$coefficients[1]
    if (m.i < 0) m.i <- mean(Data.i$time)

    Z.init[i, ] <- log(c(A.i, w.i, m.i, s.i))
  }
  Z.init
}
```

`initTheta` then sets $\boldsymbol\mu$ as the mean and $\boldsymbol\Omega$ as
the (diagonal) variance of the initial $Z^{(0)}$, and $\sigma^2$ as the mean
squared residual of the logistic fit.

```{r initTheta-fun}
initTheta <- function(Z.init, myData, omegaDiag = TRUE) {
  mu.init <- colMeans(Z.init)
  if (omegaDiag) {
    K <- ncol(Z.init)
    Omega.init <- matrix(0, K, K)
    diag(Omega.init) <- apply(Z.init, 2, var)
  } else {
    Omega.init <- var(Z.init)
  }
  res.init  <- myData$lfmc - FourParamLogis(myData$time, exp(Z.init[myData$plot, ]))
  sigma2.init <- mean(res.init^2)
  list(mu = mu.init, Omega = Omega.init, sigma2 = sigma2.init)
}
```

```{r init}
Z.init     <- initZ(myData)
theta.init <- initTheta(Z.init, myData)
tbl_Zinit <- data.frame(
  Plot  = 1:nPlot,
  Z.init
)
colnames(tbl_Zinit)[-1] <- c("log(A)", "log(w)", "log(m)", "log(s)")

knitr::kable(tbl_Zinit,
  digits    = 3,
  align     = "c",
  caption   = "Initial individual log-parameters $\\\\varphi_i^{(0)}$ obtained by per-plot optimisation."
) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE)
```

### SAEM-MCMC estimation

The SAEM algorithm is run with 100 SAEM iterations, each involving a short
MCMC exploration of the individual parameters (100 iterations), and the
Metropolis proposal scale `rho = c(1/100, 1/10, 1, 2)`. During the first
20 iterations the whole step size is used (stochastic behaviour), then it
decreases as $1/k^{0.99}$.

The SAEM step (`SAEM_MCMC_NLME`) embeds the Metropolis-within-Gibbs sampler
(`MCMC_NLME`). Full commented implementation details of the two functions are
given in [`SAEM-MCMC.R`](SAEM-MCMC.R) and discussed in **Chapter 6** of the
book.

```{r saem-params}
paramsAlgo <- list(
  nbIterMCMC = 100,
  nbIterSAEM = 100,
  rho        = c(1/100, 1/10, 1, 2),
  keep       = FALSE,
  print      = FALSE
)
paramsAlgo$gammaSAEM <- c(rep(1, 20), 1/(1:paramsAlgo$nbIterSAEM)^0.99)
```

```{r saem-estim, cache=TRUE}
res_SAEM  <- file.exists("res_SAEM_LFMC.Rdata")
res_post  <- file.exists("res_postSample_LFMC.Rdata")

if (res_SAEM) {
  # Load saved, already-processed trace of the SAEM iterations
  load("res_SAEM_LFMC.Rdata")             # -> df_res_SAEM_MCMC
} else {
  # Run the SAEM-MCMC algorithm live
  res_SAEM_MCMC <- SAEM_MCMC_NLME(myData, paramsAlgo, omegaDiag = TRUE)
  df_res_SAEM_MCMC <- as.data.frame(cbind(
    res_SAEM_MCMC$mu, res_SAEM_MCMC$sigma2))
  names(df_res_SAEM_MCMC) <- c("log(A)", "log(w)", "log(m)", "log(s)", "sigma2")
  df_res_SAEM_MCMC$iterations <- 1:(paramsAlgo$nbIterSAEM + 1)
}
```

### Trace plots of the SAEM iterations

We now plot the estimations of the population parametes obtained along the iterations. 

```{r trace-plot}
df_trace <- pivot_longer(
  df_res_SAEM_MCMC,
  cols = c("log(A)", "log(w)", "log(m)", "log(s)", "sigma2"),
  names_to = "parameter",
  values_to = "value"
)
ggplot(df_trace, aes(x = iterations, y = value)) +
  geom_line() +
  geom_vline(xintercept = 20, colour = "grey80", linewidth = 0.5) +
  facet_wrap(~ parameter, scales = "free_y", ncol = 2,
             labeller = label_parsed) +
  labs(x = "SAEM iteration", y = "Estimate")
```

The vertical line marks the end of the stochastic phase (`k = 20`). After
that, the estimates converge smoothly toward the SAEM limit.

### Point estimates

```{r estimates}
if (res_SAEM) {
  # Recompute the final theta values that were used for posterior sampling
  n <- paramsAlgo$nbIterSAEM + 1
  mu <- c(
    `log(A)` = df_res_SAEM_MCMC$`log(A)`[n],
    `log(w)` = df_res_SAEM_MCMC$`log(w)`[n],
    `log(m)` = df_res_SAEM_MCMC$`log(m)`[n],
    `log(s)` = df_res_SAEM_MCMC$`log(s)`[n]
  )
  theta.estim <- list(mu = mu, sigma2 = df_res_SAEM_MCMC$sigma2[n])
} else {
  n <- paramsAlgo$nbIterSAEM + 1
  theta.estim <- list(
    mu     = res_SAEM_MCMC$mu[n, ],
    sigma2 = res_SAEM_MCMC$sigma2[n],
    Omega  = res_SAEM_MCMC$Omega[n, , ]
  )
}

tbl_estim <- data.frame(
  Parameter = c("log(A)", "log(w)", "log(m)", "log(s)", "sigma^2"),
  Estimate  = unname(c(theta.estim$mu, theta.estim$sigma2))
)

knitr::kable(tbl_estim,
  col.names = c("Parameter", "SAEM-MCMC estimate"),
  digits    = 3,
  align     = "lc",
  caption   = "SAEM-MCMC point estimates of the population parameters."
) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE)
```

### Posterior individual trajectories

To assess the fit of the model, we sample the posterior distribution of the
individual parameters given the estimated $\widehat{\theta}$ with a longer MCMC run
(5000 iterations, keeping the trace), then compute the predicted LFMC 
trajectories with their pointwise 95% credible intervals.

```{r post-params}
paramsAlgo2 <- paramsAlgo
paramsAlgo2$keep <- TRUE
paramsAlgo2$nbIterMCMC <- 5000
paramsAlgo2$print <- TRUE
```

```{r post-sampling, cache=TRUE}
if (res_post) {
  load("res_postSample_LFMC.Rdata")      # -> post_Ind_param_MCMC
} else {
  post_Ind_param_MCMC <- MCMC_NLME(res_SAEM_MCMC$Z, myData,
                                   theta.estim, paramsAlgo2)
}
#dim(post_Ind_param_MCMC)
```

```{r post-trajectories}
seq_iter <- seq(paramsAlgo2$nbIterMCMC / 2, paramsAlgo2$nbIterMCMC, by = 2)

gg <- ggplot()
for (i in 1:nPlot) {
  Data.i <- myData |> filter(plot == i)
  seq_time <- seq(0, max(Data.i$time) + 1, length.out = 100)

  Z.post.i <- post_Ind_param_MCMC[seq_iter, i, ]
  F.i <- sapply(1:nrow(Z.post.i), function(m) {
    FourParamLogis(seq_time, exp(Z.post.i[m, ]))
  })
  q.i <- apply(F.i, 1, quantile, probs = c(0.025, 0.975), na.rm = TRUE)

  Data_simpost.i <- data.frame(time = seq_time)
  Data_simpost.i$plot_fact <- factor(i, levels = ord)
  Data_simpost.i$q05 <- q.i[1, ]
  Data_simpost.i$q95 <- q.i[2, ]

  gg <- gg +
    geom_ribbon(data = Data_simpost.i,
                aes(x = time, ymin = q05, ymax = q95,
                    fill = plot_fact, colour = plot_fact),
                alpha = 0.20, linewidth = 0.4) +
    geom_point(data = Data.i,
               aes(x = time, y = lfmc, colour = plot_fact),
               size = 2)
}

gg +
  scale_colour_manual(values = cols) +
  scale_fill_manual(values = cols) +
  labs(x = "Time (days)", y = "LFMC (%)", colour = "Plot", fill = "Plot")
```

The shaded bands are the pointwise 95% credible intervals of the predicted
LFMC trajectories for each plot, and the points are the observations. The
model reproduces the declining LFMC dynamics of the three Eastern grasses
plots well.
